---
title: "Predicting Pokemon primary type from battle statistics"
author: "Aono Li"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.path = "figures/", dpi = 150)
```

```{r libraries}
library(tidyverse)
library(tidymodels)
library(janitor)
library(corrplot)
library(glmnet)
library(ranger)
tidymodels_prefer()

set.seed(123)
```

## 1. Cleaning variable names

```{r q1-clean}
pokemon_raw <- read_csv("data/Pokemon.csv")

names(pokemon_raw)

pokemon <- clean_names(pokemon_raw)

names(pokemon)
```

`clean_names()` rewrites every column name into a consistent snake_case form: it
lowercases letters, replaces spaces and periods with underscores, and translates
punctuation into words or drops it. Here `Sp. Atk` becomes `sp_atk`, `Type 1`
becomes `type_1`, and the `#` column becomes `number`.

This matters because the original names require backtick-quoting to reference in
R and break `dplyr`, `recipes`, and formula syntax. Standardizing them once at
the start makes every downstream reference predictable and typo-resistant.

## 2. Outcome distribution and class lumping

```{r q2-barchart}
pokemon %>%
  ggplot(aes(x = fct_infreq(type_1))) +
  geom_bar() +
  labs(x = "Primary type", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

pokemon %>% count(type_1, sort = TRUE)
```

There are 18 classes of the outcome, and they are strongly unbalanced. Water
is the most common with 112 Pokémon, followed by Normal (98), Grass (70), Bug
(69), Psychic (57), and Fire (52). At the other end, Flying has only 4
Pokémon, Fairy 17, and Ice 24 — far too few to estimate a class-specific decision
boundary from, especially once the data are split into training, testing, and
five folds.

```{r q2-lump}
pokemon <- pokemon %>%
  mutate(type_1     = fct_lump_n(factor(type_1), n = 6),
         legendary  = factor(legendary),
         generation = factor(generation))

levels(pokemon$type_1)
count(pokemon, type_1, sort = TRUE)
```

`fct_lump_n(n = 6)` keeps the six most frequent levels and collapses the
remaining twelve into `Other`, leaving seven classes. Note that the pooled
`Other` category is now the largest class in the data, which is worth keeping in
mind when interpreting the model's predictions later.

## 3. Splitting and cross-validation

```{r q3-split}
pokemon_split <- initial_split(pokemon, prop = 0.80, strata = type_1)
pokemon_train <- training(pokemon_split)
pokemon_test  <- testing(pokemon_split)

dim(pokemon_train); dim(pokemon_test)

pokemon_folds <- vfold_cv(pokemon_train, v = 5, strata = type_1)
pokemon_folds
```

Stratifying the folds keeps the class proportions of `type_1` roughly constant
across all five assessment sets. Without it, the rarer types could be badly
under-represented — or absent entirely — in some folds purely by chance, which
makes the fold-level ROC AUC unstable and inflates its standard error. Since the
classes remain unbalanced even after lumping, stratification is what allows each
fold to serve as a fair miniature of the training set.

## 4. Correlation among the battle statistics

```{r q4-corrplot}
pokemon_train %>%
  select(total, hp, attack, defense, sp_atk, sp_def, speed) %>%
  cor() %>%
  corrplot(method = "number", type = "lower", diag = FALSE)
```

The categorical variables are excluded here. `type_1` and `legendary` are
unordered, so a Pearson correlation computed on their numeric codes would carry
no meaning; `generation` is ordered but its spacing is arbitrary, and correlating
it would only measure a linear trend across game releases with no natural
interpretation. The battle statistics are continuous and comparable, so the
matrix is restricted to those.

Every statistic correlates positively with `total`, from 0.57 to 0.74, but this
is definitional rather than empirical: `total` is the sum of the six components,
so it must move with each. That is also why it is excluded from the recipe in Q5.

Among the components themselves the correlations are modest and uniformly
positive, from about 0.17 to 0.51. The strongest pairs are `defense`–`sp_def`
(0.51) and `sp_atk`–`sp_def` (0.49), suggesting a general defensive axis.
`speed` is essentially uncorrelated with `defense` and only weakly related to
`hp` (0.17) — the familiar contrast between fast fragile and slow bulky Pokémon,
appearing as an absence of correlation rather than a negative one. No pair is
negatively correlated, so stronger Pokémon tend to be stronger across the board
rather than trading one stat against another.

## 5. Recipe

```{r q5-recipe}
pokemon_recipe <- recipe(type_1 ~ legendary + generation + sp_atk + attack +
                           speed + defense + hp + sp_def,
                         data = pokemon_train) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_center(all_predictors()) %>%
  step_scale(all_predictors())

pokemon_recipe
```

## 6. Elastic net specification

```{r q6-elastic-net}
en_model <- multinom_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

en_wf <- workflow() %>%
  add_model(en_model) %>%
  add_recipe(pokemon_recipe)

en_grid <- grid_regular(
  penalty(range = c(0.01, 3), trans = scales::identity_trans()),
  mixture(range = c(0, 1)),
  levels = 10
)

nrow(en_grid)
head(en_grid)
```

`penalty()` defaults to a base-10 log scale, so `trans = identity_trans()` is
required for the range 0.01–3 to be read as raw penalty values rather than as
exponents.

## 7. Random forest specification

```{r q7-random-forest}
rf_model <- rand_forest(mtry = tune(), trees = tune(), min_n = tune()) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

rf_wf <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(pokemon_recipe)

rf_grid <- grid_regular(
  mtry(range = c(1, 8)),
  trees(range = c(200, 800)),
  min_n(range = c(2, 20)),
  levels = 8
)

nrow(rf_grid)
```

`mtry` is the number of predictors randomly sampled as candidates at each split.
Restricting the candidate set is what decorrelates the trees: if every split
could consider every predictor, the strongest one would head almost every tree
and the ensemble would gain little from averaging.

`trees` is the number of trees grown in the forest. Predictions are aggregated by
majority vote across them, so more trees reduce variance, with the error curve
flattening once enough have been grown.

`min_n` is the minimum number of observations a node must contain to be split
further. Larger values stop growth earlier and produce shallower, more heavily
regularized trees; smaller values let trees grow until nodes are nearly pure.

`mtry` cannot be smaller than 1 because every split must consider at least one
candidate predictor — with zero candidates there is nothing to split on. It
cannot exceed 8 because the recipe supplies only eight predictors, and sampling
more columns than exist is undefined.

`mtry = 8` makes every predictor a candidate at every split, so no decorrelation
occurs. That is bagging — bootstrap-aggregated trees — rather than a random
forest.

## 8. Tuning

The random forest grid contains `r nrow(rf_grid)` combinations across 5 folds,
which takes far too long to run at knit time. It is tuned in the separate script
`tune_rf.R` and the results are read back in here.

```{r q8-tune-en}
en_tune <- tune_grid(
  en_wf,
  resamples = pokemon_folds,
  grid = en_grid,
  metrics = metric_set(roc_auc)
)

write_rds(en_tune, "cache/en_tune.rds")
```

```{r q8-load-rf}
rf_tune <- read_rds("cache/rf_tune.rds")
```

```{r q8-autoplot}
autoplot(en_tune)
autoplot(rf_tune)
```

**Elastic net.** Smaller penalties are uniformly better. AUC peaks at the
smallest grid value (0.01) and falls monotonically; past a penalty of about 1,
every model with a substantial lasso component has collapsed to 0.5, since pure
$L_1$ shrinkage zeroes all coefficients and the model predicts the base rate.
`mixture` matters only through how quickly that collapse arrives — ridge
(`mixture` = 0) degrades most gently and still reaches 0.62 at penalty = 3, while
mixtures above roughly 0.3 are indistinguishable and die quickly. The optimum
sits at the boundary of the prescribed range, so it has not been bracketed.

**Random forest.** `mtry` is the only hyperparameter that matters much. `mtry` =
1 is clearly worst at about 0.685; performance jumps to roughly 0.715 at `mtry` =
2–4, then drifts down toward 8. This is decorrelation at work: sampling few
predictors per split makes trees more independent, whereas `mtry` = 8 is bagging
and lets the same strong stats head every tree. `trees` and `min_n` have almost
no effect — all eight tree counts overlap within noise in every panel, and the
panels from `min_n` = 2 to 20 are nearly identical.

```{r q8-best}
collect_metrics(en_tune) %>% filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>% head(5)

collect_metrics(rf_tune) %>% filter(.metric == "roc_auc") %>%
  arrange(desc(mean)) %>% head(5)

best_en <- select_best(en_tune, metric = "roc_auc")
best_rf <- select_best(rf_tune, metric = "roc_auc")

best_en
best_rf
```

The best elastic net is `penalty` = 0.01 with `mixture` = 0.444, at a
cross-validated ROC AUC of 0.6866 (SE 0.0130). The best random forest is
`mtry` = 2, `trees` = 285, `min_n` = 2, at 0.7204 (SE 0.0130). The forest
leads by 0.034 — about 2.6 standard errors — so the nonlinear model is genuinely
better here rather than nominally ahead.

## 9. Final fit and test-set evaluation

```{r q9-final-fit}
final_rf_wf  <- finalize_workflow(rf_wf, best_rf)
final_rf_fit <- fit(final_rf_wf, data = pokemon_train)
```

```{r q9-vip}
final_rf_fit %>%
  extract_fit_parsnip() %>%
  pluck("fit", "variable.importance") %>%
  enframe(name = "variable", value = "importance") %>%
  ggplot(aes(x = importance, y = fct_reorder(variable, importance))) +
  geom_col() +
  labs(x = "Importance (impurity)", y = NULL,
       title = "Random forest variable importance")
```

The six battle statistics dominate and are nearly interchangeable: `sp_atk` is
highest at about 52, with `speed`, `attack`, `defense`, `hp`, and `sp_def` in a
tight band from roughly 45 to 50. The `generation` dummies contribute about 10
apiece, and `legendary_TRUE` is least useful at about 5.

This is broadly expected. Primary type is a design property that the stat
distribution reflects, so the continuous stats carry the signal. Generation is
arbitrary with respect to type — every game introduces Pokémon of every type —
and legendary status is rare and spread across types, so neither should help.
Mildly surprising is how evenly the six stats rank; part of that is an artifact
of impurity importance, which favors continuous predictors with many possible
split points over binary dummies.

```{r q9-test-preds}
rf_preds <- augment(final_rf_fit, new_data = pokemon_test)

rf_preds %>%
  roc_auc(truth = type_1,
          .pred_Bug, .pred_Fire, .pred_Grass, .pred_Normal,
          .pred_Psychic, .pred_Water, .pred_Other)
```

```{r q9-roc-curves}
rf_preds %>%
  roc_curve(truth = type_1,
            .pred_Bug, .pred_Fire, .pred_Grass, .pred_Normal,
            .pred_Psychic, .pred_Water, .pred_Other) %>%
  autoplot()
```

```{r q9-confusion}
rf_preds %>%
  conf_mat(truth = type_1, estimate = .pred_class) %>%
  autoplot(type = "heatmap")
```

```{r q9-accuracy}
rf_preds %>% accuracy(truth = type_1, estimate = .pred_class)
```

## 10. What the model does and does not do

Overall the model is mediocre: multiclass ROC AUC of 0.6816 on the testing
set and accuracy of 0.410. The testing AUC falls 0.039 below the
cross-validated 0.7204 — roughly three fold-level standard errors, a real drop
rather than resampling noise. Two things contribute: the testing set holds only
161 observations, so a single estimate is volatile; and `select_best()` takes the
maximum over 512 configurations, which is optimistically biased, so part of the
winning fold-average was luck.

The confusion matrix and the ROC curves tell different stories. On hard
predictions the model is good at `Other` (62 of 70 correct) and close to useless
elsewhere — Fire, Grass, and Water receive no correct predictions at all, and
Bug, Normal, and Psychic get one or two each. That reflects imbalance more than
modeling: `Other` pools twelve types into 342 of 800 Pokémon, so the argmax rule
sends nearly everything there.

The per-class ROC curves assess ranking rather than the argmax decision, and are
more informative. Bug is clearly best separated, with Fire and Normal also well
above the diagonal; Psychic is moderate; Grass sits close to the diagonal and
Water is essentially on it. So the model can rank Pokémon by their probability of
being Bug even though it almost never predicts Bug outright — Bug Pokémon cluster
at low, distinctive stat totals. Water is hardest: it is the largest real type
and its members span nearly the whole stat range, from Magikarp to legendaries,
leaving no signature to learn. Grass is similar. The broader lesson is that
primary type is only weakly determined by battle statistics, and the model
succeeds only where a type occupies a distinctive corner of stat space.
